Modal Verification Patterns for Systems

### 2025-06-16 paper.tex
+

\subsection{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification. 
\paragraph{Modality for Recursive Modal Contexts on Trees}
\label{sec:overviewnode}

@@

%\setlength\intextsep{0pt}
\begin{wrapfigure}{r}{0.25\textwidth}
  \centering  \includegraphics[width=0.25\textwidth]{entail.png}
  \caption{Understanding the level of $\varphi$ satisfaction in the \textsc{CoW} tree of modal nodes}
   \label{fig:nodeentail} 
\end{wrapfigure}
We use a modality similar to $\text{\faSitemap}@[\ell]\phi$ to indicate that $\phi$ is true of the tree node rooted at $\ell$, as well as at \emph{all reachable children} of that node.
This means that establishing a particular $\phi$ throughout a tree must be done in a bottom-up manner (see Figure \ref{fig:nodeentail}).
Capturing this modally, rather than with more typical ad-hoc hand-crafted invariants, offers advantages in a number of proof steps, in addition to concisely capturing intuitive reasoning about subtrees with all nodes satisfying a particular property. One basic example is that while the actual disk sync code requires the full filesystem tree to be in an amplified state (hence write amplification), most of the filesystem works with more general trees. Since $\phi_{\textsf{amplified}}\vdash\phi_{\textsf{valid}}$, using our \faSitemap modality we can conclude $\text{\faSitemap}@[\ell]\phi_{\textsf{amplified}}\vdash\text{\faSitemap}@[\ell]\phi_{\textsf{valid}}$ by an analogue of Axiom K.
There are also other places in the code where a subtree satisfying a stronger assertion at all nodes must be linked into a larger tree satisfying only a weaker assertion overall: predicate representing the valid tree as the tree of nodes which are backed by resources allocated pages, i.e. $\phi_{\textsf{valid}}$, is weaker than the predicate representing the tree of nodes being currently amplified in the current snapshot. Again, the same reasoning principle makes this straightforward. As alluded to earlier, our $\phi_{\textsf{amplified}}$ and company are in fact \emph{predicates} parameterized by a choice of node, though by lifing \textsf{Iris}'s BI structure they can use all of the standard separation logic connectives. The rest of this subsection develops this modality in full detail.
%\colinsaysinl{I cut some text here about the three-way split (commented out) that I wasn't sure how to use, but maybe I deleted something important?}
%To put it concretely, as shown in Figure \ref{fig:cow_index_tree.JPG}, \textsc{CoW} semantics determine what we can claim on the structure of the filesystem indexing tree:
%\begin{itemize}
%\item The bottom-up direction splits the tree into the two half at the cursor node: 1. the bottom part accessible from the cursor node and per-node-amplification is reflected in case needed 2. the upper part on which amplifcation has not reflected yet. 
%\item Horizontally, amplifcation path splits the bottom part of the tree into three different contexts: untouched tree nodes, nodes on the amplification path, and the ones generated as part of node splitting. 
%\end{itemize}
%Based on these two observations, in fact, bottom-up amplification determines the choice of modal context by giving the boundries of \textit{any claimable knowledge} (e.g. $\varphi_{\textsf{amplified}}$, $\varphi_{\textsf{untouched}}$): a subtree bounded by the cursor node.
\subsection{Speculation in Consistency}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\loc \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. We adapt these ideas to assertions of the form $@[l]\varphi$asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Shallow Modalities in Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, which for \textsf{Iris} would be a collection of \textsc{CMRA}s; this is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{krebbers2017essence}. However in general, many modalities can be given proof-theoretic semantics directly within a logic. This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} logic itself), as well as for a wide range of other modalities in the literature~\cite{restall2002introduction}. These modalities are typically multi-modal, with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ is typically replaced with substructural implication $\sepimp$.

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. This will allow our modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\subsubsection{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion $\mathsf{minCommitIndex}(5)$ to represent the former knowledge,
one would need to separately provide custom reasoning to conclude $\mathsf{minCommitIndex}(3)$.
\section{Contingency Decomposition of a System}
\todo[inline]{Colin says: I don't think this section needs to change much,
but it should be read carefully with the new audience in mind,
possibly giving more details on what systems concepts each system is dealing with, for a non-systems audience.}

>>>>

\subsection{Virtualization}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from a system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.

\subsection{Speculation in Consistency}
%\subsection{Sharing} Sharing is one of the fundamental reasons for kernels to {virtualize} resources,
%for creating the illusion of exclusive access to virtualized versions of shared resources, including CPU time (via preemption).
%An implementation of sharing relies on two concepts: \emph{indirection} and \emph{coordination}. Indirection is a method that allows the mapping of resources of one type into another. By doing so, it allows treating the limited resources (e.g. hardware resources such as physical memory) as if there are more than they actually are (e.g. via virtual memory addresses and paging). Consequently, virtualization for the sake of ensuring availability requires \emph{sharing} of the resources that exist in reality.
%%\paragraph{Isolation Hand-in-Hand with Sharing}
%%\Ismail{put a few words on abstractions such as address space, stack region, snapshots ,etc.}
%%\Colin{Not sure what you want here?}

\paragraph{Hybrid Logic} Hybrid logics~\cite{blackburn1995hybrid,goranko1996hierarchies,gargov1993modal,areces2001hybrid,brauner2010hybrid} are a class of modal logics that add two primary new concepts to a logic. Nominals $\loc \in \textsf{Loc}$ uniquely identify points in a model (e.g., a particular state), so there is exactly one point in the model where a nominal $\loc$ is true. Satisfaction operators are modal operators indexed by nominals, which enable claims about the truth of another proposition at some arbitrary point in the model identified by a nominal. Traditionally $@_l\varphi$ asserts that $\varphi$ is true in the state identified by $l$. Conceptually, given our context, we might like $l$ to indicate truth in a particular data structure $l$.  However, most prior work on hybrid logics works in classical logics rather than a substructural setting like separation logic. We adapt these ideas to assertions of the form $@[l]\varphi$asserts that $\varphi$ is true in a data structure with \emph{logical} state $\sigma$ \emph{at location} $l$.
In practice we must design variants of this for each kind of structure we wish to apply this to, e.g. tree nodes, free list, or snapshots.
%This section provides details of and models for each modal abstraction, focusing on how each data structure can act as a model for a data-structure-specific logic.
%We continue to motivate the development with aspects of the write amplification process described in the previous section; Section \ref{sec:evaluation} provides further examples using the modal abstractions in proofs.
\paragraph{Temporal Operators}
The best known class of modal logics in the systems community is undoubtedly temporal logics (whether \textsc{LTL}~\cite{pnueli1977temporal}, \textsc{CTL}~\cite{emerson1982using}, \textsc{TLA}~\cite{lamport1994temporal}, or others),
due in part to Lamport's influence~\cite{lamport2002specifying} leading to its not-infrequent use in specifying
distributed algorithms \cite{ongaro2014search,ailijiang2019wpaxos}.
\paragraph{Nominalization}
Some classes of assertions benefit specifically from \emph{naming} the explicit conditions where they are true (as opposed to simply requiring them to be true \emph{somewhere} or \emph{everywhere} as in the most classic modal operators).
This naming generally resembles
the \emph{satisfaction} operator of 
\emph{hybrid logic}~\cite{blackburn1995hybrid,areces2001hybrid}: $@_\iota P$ which evaluates the truth of $P$ at the named (Kripke model) state $\iota$. For this reason we refer to the general idea of naming circumstances explicitly as \emph{nominalization},
even though the examples we discuss are not necessarily actually hybrid logics.

In each of these cases, the fact that these facts are described using modalities with the core modal property $P \Rightarrow Q$, then $M(P) \Rightarrow M(Q)$ means, for example, that if a
process $p$ knows that the commit index is greater than 5 (e.g., $\mathcal{K}_p(\mathsf{commitIndex} > 5)$)
no extra work is required to conclude that the node knows it is greater than 3 (i.e., that $\mathcal{K}_p(\mathsf{commitIndex} > 3)$),
because this follows from standard properties of modal operators as described above.
In contrast, if verification instead used a custom assertion $\mathsf{minCommitIndex}(5)$ to represent the former knowledge,
one would need to separately provide custom reasoning to conclude $\mathsf{minCommitIndex}(3)$.
\paragraph{Shallow Modalities in Iris Separation Logic}
\label{sec:modal_in_iris}
Classically modal operators are given direct semantics in terms of an underlying model, which for \textsf{Iris} would be a collection of \textsc{CMRA}s; this is the case for \textsf{Iris}'s one built-in modality for step-indexing~\cite{krebbers2017essence}. However in general, many modalities can be given proof-theoretic semantics directly within a logic. This is the case for \textsf{Iris}'s weakest precondition modality (which is defined within the \textsf{Iris} logic itself), as well as for a wide range of other modalities in the literature~\cite{restall2002introduction}. These modalities are typically multi-modal, with a form of roughly $M~\overline{ctxt}~\phi\triangleq\ldots\rightarrow \phi~\overline{ctxt}$. In a substructural setting~\cite{restall1993modalities,krebbers2017essence} the implication $\rightarrow$ is typically replaced with substructural implication $\sepimp$.

In addition, the modal definitions themselves may represent ownership of resources in a substructural setting~\cite{restall1993modalities,dovsen1992modal,kamide2002kripke}. This will allow our modalities to simultaneously represent ownership of certain resources used for interpreting the modality itself, as well as whatever resource ownership is implied by the modal argument.
\paragraph{Modal Abstractions for Recursive Trees Structure with Nominals}
\label{sec:overviewnode}

@@

%\setlength\intextsep{0pt}
\begin{wrapfigure}{r}{0.25\textwidth}
  \centering  \includegraphics[width=0.25\textwidth]{entail.png}
  \caption{Understanding the level of $\varphi$ satisfaction in the \textsc{CoW} tree of modal nodes}
   \label{fig:nodeentail} 
\end{wrapfigure}
We use a modality similar to $\text{\faSitemap}@[\ell]\phi$ to indicate that $\phi$ is true of the tree node rooted at $\ell$, as well as at \emph{all reachable children} of that node.
This means that establishing a particular $\phi$ throughout a tree must be done in a bottom-up manner (see Figure \ref{fig:nodeentail}).
Capturing this modally, rather than with more typical ad-hoc hand-crafted invariants, offers advantages in a number of proof steps, in addition to concisely capturing intuitive reasoning about subtrees with all nodes satisfying a particular property. One basic example is that while the actual disk sync code requires the full filesystem tree to be in an amplified state (hence write amplification), most of the filesystem works with more general trees. Since $\phi_{\textsf{amplified}}\vdash\phi_{\textsf{valid}}$, using our \faSitemap modality we can conclude $\text{\faSitemap}@[\ell]\phi_{\textsf{amplified}}\vdash\text{\faSitemap}@[\ell]\phi_{\textsf{valid}}$ by an analogue of Axiom K.
There are also other places in the code where a subtree satisfying a stronger assertion at all nodes must be linked into a larger tree satisfying only a weaker assertion overall: predicate representing the valid tree as the tree of nodes which are backed by resources allocated pages, i.e. $\phi_{\textsf{valid}}$, is weaker than the predicate representing the tree of nodes being currently amplified in the current snapshot. Again, the same reasoning principle makes this straightforward. As alluded to earlier, our $\phi_{\textsf{amplified}}$ and company are in fact \emph{predicates} parameterized by a choice of node, though by lifing \textsf{Iris}'s BI structure they can use all of the standard separation logic connectives. The rest of this subsection develops this modality in full detail.
%\colinsaysinl{I cut some text here about the three-way split (commented out) that I wasn't sure how to use, but maybe I deleted something important?}
%To put it concretely, as shown in Figure \ref{fig:cow_index_tree.JPG}, \textsc{CoW} semantics determine what we can claim on the structure of the filesystem indexing tree:
%\begin{itemize}
%\item The bottom-up direction splits the tree into the two half at the cursor node: 1. the bottom part accessible from the cursor node and per-node-amplification is reflected in case needed 2. the upper part on which amplifcation has not reflected yet. 
%\item Horizontally, amplifcation path splits the bottom part of the tree into three different contexts: untouched tree nodes, nodes on the amplification path, and the ones generated as part of node splitting. 
%\end{itemize}
%Based on these two observations, in fact, bottom-up amplification determines the choice of modal context by giving the boundries of \textit{any claimable knowledge} (e.g. $\varphi_{\textsf{amplified}}$, $\varphi_{\textsf{untouched}}$): a subtree bounded by the cursor node.

\section{Contingency Decomposition of a System}
\todo[inline]{Colin says: I don't think this section needs to change much,
but it should be read carefully with the new audience in mind,
possibly giving more details on what systems concepts each system is dealing with, for a non-systems audience.}
